Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-19 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Getting close

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season \citep{Nelson-2001}.

@@

These species-specific flammability effects are increasingly well established.
However, how moisture dynamics modulate species specific flammability is poorly
known. Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
to variation in fire regime across landscapes \citep{Stephens-2001}.

@@

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
estimates based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these vary in relation to specific leaf area and litter packing density.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. We tested if the nonadditive effects on reported in
previous studies occur under higher moisture levels, and whether fuel moisture
effects strengthen or moderate any nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed or lightly decomposed). Leaf litter was
collected in the summer (mid-June to mid-July) in 2012 from 21 separate sites
across the study area. We collected from a minimum of four populations per
species with each population within a species was 9--32\,km distant from its
neighbors. An exception to this was \emph{A. magnifica} which is restricted to
the higher elevations and could only be collected from one area due to
designated wilderness boundaries and logistical constraints. Within that area,
\emph{A. magnifica} litter was collected from over 10 different individuals.
For the remaining species, the collection involved 2--4 individual trees at
each site which were separated from one another by at least 10\,m. The litter
was collected 2\,m away from the trunk to obtain a more uniform sample of leaf
and fine branchlet litter, because bark and heavier branches tend to fall
closer to the tree trunk. The top 2.5-5\,cm of litter were collected by raking
and shoveling into large bags. All litter was oven dried to $< 5$\% fuel
moisture by dry weight, then weighed and divided into moisture trials. The
collected litter was stored in large paper bags and cardboard boxes in a lab
storage space. We minimized handling and moving the bagged litter to avoid
mechanical damage and changes to the particle size distributions. The moisture
manipulations and subsequent burning trials were conducted in April--August
2015.

@@

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and of differences in
drying rate. This has consequences for flammability as a function of species
and time since wetting. Our flammability trials demonstrate that both flame
spread rate and fuel consumption increased with time since wetting but that the
rate of increase varied across these taxa. Only the pines are able to ignite
and support flame spread immediately after wetting (Fig.
\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
response to drying with no ignition prior to 72 hours of drying but once
ignition is possible, it had faster spread rates than the other taxa. The
Cupressaceae, which in our two species have dense litter but low maximum water
absorption had low flammability across moisture levels. This result is
consistent with the low flammability found in an Australian Cupressaceae that
have small litter particles and dense packing \citep{Scarff+Westoby-2006,
  Bowman+Haverkamp+etal-2018}. However, the densely packed litter is not a
consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
found relatively high litter flammability and low litter density for some Asian
Cupressaceae.

>>>>

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season.

@@

These species-specific flammability effects are increasingly well established.
However, how moisture dynamics modulate species specific flammability is poorly
known. Fuel moisture often limits fire activity in the western US
\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists often
categorize fuel by size classes that correspond to moisture exchange rates,
although such simplified categories have long been recognized as not sufficient
to describe species-specific effects on moisture dynamics
\citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon
forest structure (shading and rain throughfall effects), litter physical
structure, and chemical composition of the fuel-beds \citep{Nelson+Hiers-2008,
  Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing
moisture dynamics of fine fuels are used by fire scientists \citep{Viney-1991,
  Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to
describe equilibrium conditions and at least include temperature and relative
humidity although some include precipitation and solar radiation effects as
well. Explicit models of fire behavior as a function of time since wetting
could improve fire prediction, however. Drying rates following snowmelt or
following rainfall during the fire season will influence the length of the fire
season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Differential drying can be especially
important when precipitation can occur during the same season as ignitions.
Although synoptic weather conditions that create dry and wet lightning storms
differ, the potential for precipitation and lightning during the same overall
period of thunderstorm activity often exists: wet and dry convection days can
occur under overlapping conditions, \citep{Rorig+Ferguson-1999}. Moisture
levels of fine fuels respond rapidly to precipitation events during the fire
season \citep{Estes+Knapp+etal-2012}. Differential moisture content across time
may contribute to the often high spatial heterogeneity of surface fires
\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
variation in fire regime across landscapes \citep{Stephens-2001}.

@@

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
flammability measurements based on dry fuels. We investigated these litter
moisture dynamics across eight tree species in a temperate fire-prone forest in
California. We determined the maximum water absorption and rate of moisture
loss and examined how these traits vary in relation to specific leaf area and
litter packing density. Experimental burn trials of leaf litter at various
moisture levels determined how time since wetting influenced two axes of
flammability characterized by flame spread rate and total fuel consumed.
Previous work described the flammability of these species in monocultures and
in mixtures for oven-dry litter \citep{Magalhaes+Schwilk-2012} and found
consistent positive nonadditivity of flammability where fire behavior of
mixtures was driven by the most flammable species. We tested if the nonadditive
effects on reported in previous studies occur under higher moisture levels, and
whether fuel moisture effects strengthen or moderate any nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried to $< 5$\% fuel moisture by dry weight, then weighed
and divided into moisture trials. The collected litter was stored in large
paper bags and cardboard boxes in a lab storage space. We minimized handling
and moving the bagged litter to avoid mechanical damage and changes to the
particle size distributions. The moisture manipulations and subsequent burning
trials were conducted in April--August 2015.

@@

The dry down curves we measured demonstrate that after a rain event, the litter
of different species will exhibit different moisture contents over time as a
consequence of both differential initial absorption and of differences in
drying rate. This has consequences for flammability as a function of species
and time since wetting. Our flammability trials demonstrate that both flame
spread rate and fuel consumption increased with time since wetting but that the
rate of increase varied across these taxa. Only the pines are able to ignite
and support flame spread immediately after wetting (Fig.
\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
response to drying with no ignition prior to 72 hours of drying but once
ignition is possible, it had faster spread rates than the other taxa. Our two 
Cupressaceae, which have dense litter and low maximum water
absorption, had low flammability across moisture levels. This result is
consistent with the low flammability found in an Australian Cupressaceae that
have small litter particles and dense packing \citep{Scarff+Westoby-2006,
  Bowman+Haverkamp+etal-2018}. However, the densely packed litter is not a
consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
found relatively high litter flammability and low litter density for some Asian
Cupressaceae.
